---
title: "Working with data"
output: html_document
---

<!-- This file by Matt Nickodemus is licensed under a Creative Commons Attribution 4.0 International License -->

# The basics of ggplot2

Every layer in a ggplot has five components

- data
- aesthetics
- geom
- stat
- position

In addition to these five components we need to understand scales, guides, and themes. We will do this with a set of examples. Our first example is straight forward line plot showing the number of credits taken in each college from 2011 to 2020. As we walk though this, we will look at six components

- data 
- aesthetics
- geom

and 

- scales
- guides 
- themes



## Creating our data set

We will create a data set consisting of college credit counts for 2011-2020.

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(purrr)

make_credits <- function(college) {
  slope <- rnorm(1, 500, 500)
  intercept <- rnorm(1, 20000, 2000) %>% floor()
  x <- 1:10
  college <- rep(college, 10)
  noise <- rnorm(10, 1000, 1000)
  
  credits <- intercept + slope*x + noise
  
  output_df <- tibble(college = college, year = 2010 + x, credits = credits ) %>% 
    mutate(credits = as.integer(credits))
  
  return(output_df)
}


colleges <- c('Fine Art', 'Business', 'Physical Science', 'Social Science', 
              'Health Science', 'Humanities', 'Life Science', 'Earl')

credits_by_college <- map_dfr(colleges, make_credits)
```

This data output is in tidy format. In tidy data 

- Each variable forms a column.
- Each observation forms a row.
- Each type of observational unit forms a table.

For our data set, the observations are the number of credits offered by a specific
college in a given term. The variables are year and credits. Your data should always be tidy when using ggplot.

```{r, messages = FALSE}
library(gt)
head(credits_by_college) %>% gt()
```

# Creating a plot

Now we will create our plot. We will begin with the data aesthetic.

## data
In our plot the data layer is created with the following code. Note that there is nothing on the plot. There are no axes, no graphs, no colors, no legend. Just a blank canvas. The reason for this is that in ggplot, the plot is created when we map the variables in the data to visual elements in the plot. This is done with the aes() function. 

```{r}
ggplot(credits_by_college)
```

## aesthetics

When we add the `aes()` call within the `ggplot` call, we produce a set of axes on our plot. This is because we have mapped the variable year to the x axis in our graph, and the variable credits to the y-axis in our graph. We do not see a plot yet because we have not yet told ggplot what type of geometric object we want to use to in our plot. We do that next.

```{r}
ggplot(credits_by_college, aes(x = year, y = credits))
```


## geom

### Huh?

In our next plot we have chosen the line geom for our graph. This does not give us the result that we are looking for. We are trying to draw a line chart showing credits for each year, for each college. The mistake we have made is that we have not assigned the college variable to a visual aspect of our graph. We do so in the next graph.

```{r}
ggplot(credits_by_college, aes(x = year, y = credits)) + geom_line()
```

### More aesthetics?

This graph gives us something closer to what we are looking for, but it is still not quite there.

```{r}
ggplot(credits_by_college, aes(x = year, y = credits, group = college)) + geom_line()
```

The problem is that we would like the lines to be different colors, and we would like each color to be associated with a color. So we have to map the variable college to another aesthetic, color. 

```{r}
ggplot(credits_by_college, aes(x = year, y = credits, group = college, color = college)) + 
  geom_line() + geom_point(size = .5, alpha = .8)
```

## Scale

```{r}
library(scales)
ggplot(credits_by_college, aes(x = year, y = credits, group = college, color = college)) + 
  geom_line() + geom_point(size = .5, alpha = .8) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma)
```


## Guide

```{r}
ggplot(credits_by_college, aes(x = year, y = credits, group = college, color = college)) + 
  geom_line() + geom_point(size = .5, alpha = .8) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma) +
  guides(color = guide_legend(title = "College"))
```


## Labels

```{r}
ggplot(credits_by_college, aes(x = year, y = credits, group = college, color = college)) + 
  geom_line() + geom_point(size = .5, alpha = .8) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma) +
  guides(color = guide_legend(title = "College")) +
  labs(title = "Credits by college",
       subtitle = "Credits taken within each college 2011-2020",
       caption = "Data supplied by OIE") +
  xlab("Year") +
  ylab('Credits')
```

## Theme

```{r}
ggplot(credits_by_college, aes(x = year, y = credits, group = college, color = college)) + 
  geom_line() + geom_point(size = .5, alpha = .8) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma) +
  guides(color = guide_legend(title = "College")) +
  labs(title = "Credits by college",
       subtitle = "Credits taken within each college 2011-2020",
       caption = "Data supplied by OIE",
       x = 'Year',
       y = 'Credits') +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.subtitle = element_text(color = "#a6a6a6", size = 10),
    plot.caption = element_text(color = '#a6a6a6', size = 8, face = 'italic')
  )
  
```

## Custom Colors

```{r}

my_colors <- c("#8E877C", "#E7CA8C", "#F6BC8C", "#C87D55", "#868662", "#AD7C82", 
               "#DC5A3F", "#668CB0")
names(my_colors) <- colleges

ggplot(credits_by_college, aes(x = year, y = credits, group = college, color = college)) + 
  geom_line() + geom_point(size = .5, alpha = .8) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma) +
  guides(color = guide_legend(title = "College")) +
  labs(title = "Credits by college",
       subtitle = "Credits taken within each college 2011-2020",
       caption = "Data supplied by OIE",
       x = 'Year',
       y = 'Credits') +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.subtitle = element_text(color = "#a6a6a6", size = 10),
    plot.caption = element_text(color = '#a6a6a6', size = 8, face = 'italic')
  ) +
  scale_colour_manual(values = my_colors)
```
