---
title: "More dplyr"
output: html_document
---

```{r setup, warning = FALSE, message = FALSE}
library(here)
library(DT)
library(gt)
source("data_io_util.R")
library(tidyverse)
library(readr)
library(scales)

knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
```

# Pull the data
```{r data pull}
student_list <- get_data_from_sql(file_name="student_list.sql", dsn="BRPT")
saveRDS(student_list, file = here::here('sensitive', 'student_list.RDS'))
student_list <- readRDS(here::here('sensitive', 'student_list.RDS'))
```

# Munge the data

Here we are going to shape the data into what we want.

## Aggregate by Advisor


```{r}
aggregate_by_advisor <- student_list %>% 
  select(advisor_full_name, student_banner_id) %>% 
  group_by(advisor_full_name) %>% 
  count() %>% 
  ungroup()
```


### Table


```{r}
aggregate_by_advisor %>% datatable()
```

### Barchart

## GROUP BY and COUNT

Now suppose that we want to count how many graduates each college had. To do this we will use the `group_by` and `mutate` functions from the `tidyverse`.



Now let's throw this into a bar chart.

 
```{r}
ggplot(aggregate_by_advisor,
       aes (x = reorder(advisor_full_name, n), y = n, fill = advisor_full_name)) + # the reorder function arranges to bars
  geom_bar(stat='identity') +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none",
        plot.subtitle = element_text(color = "#a6a6a6", size = 10),
    plot.caption = element_text(color = '#a6a6a6', size = 8, face = 'italic')
    ) +
  labs(title = "Number of students by Academic Advisor",
       subtitle = "",
       caption = "Data supplied by OIE",
       x = 'Academic Advisor',
       y = 'Number of Students')
```


# Aggregate by College

```{r}
aggregate_by_college <- student_list %>% 
  select(student_college, student_banner_id) %>% 
  group_by(student_college) %>% 
  count() %>% 
  ungroup()
```

### Table


```{r}
aggregate_by_college %>% datatable()
```

### Barchart

```{r}
ggplot(aggregate_by_college,
       aes (x = reorder(student_college, n), y = n, fill = student_college)) + # the reorder function arranges to bars
  geom_bar(stat='identity') +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none",
        plot.subtitle = element_text(color = "#a6a6a6", size = 10),
    plot.caption = element_text(color = '#a6a6a6', size = 8, face = 'italic')
    ) +
  labs(title = "Number of students by Academic Advisor",
       subtitle = "",
       caption = "Data supplied by OIE",
       x = 'College',
       y = 'Number of Students')
```


## Aggregate by Program


```{r}
aggregate_by_program <- student_list %>% 
  select(student_program, student_banner_id) %>% 
  group_by(student_program) %>% 
  count() %>% 
  ungroup()
```



### Table


```{r}
aggregate_by_program %>% datatable()
```
### Barchart

```{r}
ggplot(aggregate_by_program,
       aes (x = reorder(student_program, n), y = n, fill = student_program)) + # the reorder function arranges to bars
  geom_bar(stat='identity') +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none",
        plot.subtitle = element_text(color = "#a6a6a6", size = 10),
    plot.caption = element_text(color = '#a6a6a6', size = 8, face = 'italic')
    ) +
  labs(title = "Number of students by Academic Advisor",
       subtitle = "",
       caption = "Data supplied by OIE",
       x = 'Program',
       y = 'Number of Students')
```